The Sub-Millimetre User Reduction Facility: an iterative map-maker for SCUBA-2

### 2012-01-27 smurf_paper.tex
more text for cosmology section

\scuba\ surveys designed to detect extremely faint point-sources
(e.g. high-redshift star-forming galaxies, and debris disks) are
ideally limited by the white-noise performance of the instrument. The
approach described here for maximizing the S/N of point-sources
involves three major steps: (i) generating a map that removes most
low-frequency noise sources with approximately linear response without
prior knowledge of the location of sources; (ii) \emph{whitening}
(essentially flattening) the map to make the noise properties of the
map easier to deal with; and (iii) detecting point sources using a
\emph{matched-filter}. Note that variations on this general procedure
have been used extensively in the submm cosmology community using
previous instruments
\citep[e.g.][]{scott2002,laurent2005,coppin2006,scott2008,perera2008,devlin2009}.
In this section we will reduce data of the Lockman Hole taken in 2010
during S2SRO as a pilot project for the Cosmology Legacy Survey. It
consists of $\sim8.5$\,hours of data taken in 36 separate scans
(average 15\,min. each) spread over February and March, covering an
area of about 50\,arcmin$^2$. ... full obs list ... dropped scans due
to terrible noise properties ...

@@

\begin{figure}
\centering
\includegraphics[width=\linewidth]{lockman_pspec.pdf}
\caption{Azimuthally-averaged angular power spectral densities for the
  raw map output by SMURF (dashed black line), the jackknife (solid
  orange line), and whitened (solid black line) maps (panels (a), (d),
  and (b) from Fig.~\ref{fig:lockman_maps}, respectively. Since the
  raw map contains spatially correlated signals on large scales both
  due to noise and astronomical signals (low-spatial frequencies), the
  jackknife map (difference of two approximately equal-length subsets
  of the data) is used to generate a plausible realization of pure
  noise. Assuming that the noise properties are isotropic, a whitening
  filter is estimated from the reciprocal of the jackknife power
  spectrum with only a radial dependence. The power spectrum of the
  resulting signal map has residual power on large scales which is
  presumably due to astronomical sources.}
\label{fig:lockman_pspec}
\end{figure}

Since the \scuba\ scan strategy is usually isotropic (all position
angles scanned with roughly equal weights), we make the simplifying
assumption that the angular noise power spectrum is azimuthally
symmetric (inspection of the 2-dimensional FFT confirms
this). ... calculate filter as slightly-smoothed complement of
azavpspec ... Finally, the filter is applied to the map, resulting in
panel...

@@

>>>>

\scuba\ surveys designed to detect extremely faint point-sources
(e.g. high-redshift star-forming galaxies, and debris disks) are
ideally limited by the white-noise performance of the instrument. The
approach described here for maximizing the S/N of point-sources
involves three major steps: (i) generating a map that removes most
low-frequency noise sources with approximately linear response,
without prior knowledge of the location of sources; (ii)
\emph{whitening} (essentially flattening) the map to make the noise
properties of the map easier to deal with; and (iii) detecting point
sources using a \emph{matched-filter}. Note that variations on this
general procedure have been used extensively in the submm cosmology
community using previous instruments
\citep[e.g.][]{scott2002,laurent2005,coppin2006,scott2008,perera2008,devlin2009}.
In this section we will reduce scans of the Lockman Hole taken in 2010
during S2SRO as a pilot project for the Cosmology Legacy Survey. It
consists of $\sim8.5$\,hours of data taken in 36 separate scans
(average 15\,min. each) spread over February and March, covering an
area of about 50\,arcmin$^2$. The full list of dates and observation
numbers includes 20100218: 63, 64, 65, 70, 71, 72, 90, 91, 92, 97, 98,
99; 20100220: 111, 112, 113, 118, 128, 129, 130; 20100303: 61, 62, 64,
69, 70, 72, 73, 74, 75; 20100309: 87, 88, 90, 91; and 20100311: 59,
64, 65, 72. These observations represent about 80\% of the total data
taken as part of the project; the remaining observations were dropped
due to problems keeping the arrays properly locked during S2SRO, and
were easily identified by their highly variable and erratic bolometer
time-series (which resulted in maps full of large streaks).

@@

\begin{figure}
\centering
\includegraphics[width=\linewidth]{lockman_pspec.pdf}
\caption{Radial (azimuthally-averaged) angular power spectral
  densities for the raw map output by SMURF (dashed black line), the
  jackknife (solid orange line), and whitened (solid black line) maps
  (panels (a), (d), and (b) from Fig.~\ref{fig:lockman_maps},
  respectively (considering only the square region indicated in
  Fig.~\ref{fig:lockman_maps}d). Since the raw map contains spatially
  correlated signals on large scales both due to noise and
  astronomical signals (low-spatial frequencies), the jackknife map
  (difference of two approximately equal-length subsets of the data)
  is used to generate a plausible realization of pure noise. Assuming
  that the noise properties are isotropic, a whitening filter is
  estimated from the reciprocal of the jackknife power spectrum with
  only a radial dependence. The power spectrum of the resulting signal
  map still has residual power on large scales, which is presumably
  due to astronomical sources.}
\label{fig:lockman_pspec}
\end{figure}

Since the \scuba\ scan strategy is usually isotropic (all position
angles scanned with roughly equal weights), we make the simplifying
assumption that the angular noise power spectrum is azimuthally
symmetric. For these data, there are no obvious anisotropic structures
in the 2-dimensional FFT. The radial (azimuthally-averaged) angular
power spectrum therefore encodes all of the useful information about
the noise properties. These power spectra for the raw output of SMURF,
and the jackknife map (transforming only the approximately uniform
region indicated by the square in Fig.~\ref{fig:lockman_maps}d in each
case) by the dashed black, and solid orange lines in
Fig.~\ref{fig:lockman_pspec}, respectively. Both power spectra are
approximately flat at spatial frequencies $\gsim 0.06$\,arcsec$^{-1}$
(scales $\lsim16$\,arcsec), with the exception of a narrow line
feature at $\sim0.175$\,arcsec$^{-1}$ (a scale of
$\sim5.7$\,arcsec). A possible theory for the feature at 5.7\,arcsec
is that it is related to the inter-bolometer spacing in the focal
plane (which happens to be this size); due to small relative drifts in
the baselines of adjacent bolometers, there may be faint parallel
stripes in the map along the scan direction (the superposition of many
scans at different angles then results in an isotropic noise
pattern). It is not likely that this signal is due to astronomical
sources because it appears with a nearly equal amplitude in both the
real map and the jackknife.  At lower spatial frequencies, both the
real map and the jackknife power spectra grow significantly, as a
result of the more obvious large-scale undulations in
Figs.~\ref{fig:lockman_maps}(a) and (d).

In this example, to identify point sources, we first filter the maps
to suppress the large-scale waves. We use the radial power spectrum of
the jackknife map (orange line in Fig.~\ref{fig:lockman_pspec}) as it
should contain no appreciable contamination from astronomical sources.

@@

default: converges after 21 iterations
